Capabilities, MMIO and Robust Safety

### 2020-04-02 capability-notes/driver-notes/main.tex
driver notes: move most of the remarks to an appendix to ease the reading of the main document

@@

Notice how in the second rule, we read an arbitrary integer $x$, and record it
in the trace.

The \I{Store} rule would be similar.

\newcommand{\stdin}{\X{in}}
\begin{remark}
As Lars remarked, the read rule introduces non-determinism into our operational
semantics.
This has lead to some complications in the past (the nature of which is still
mysterious to me; n.b. that we use\\ $\I{ wp\_lift\_atomic\_head\_step\_no\_fork}$
in our WP rules, and that one would allow non-determinism).
(one question I -Thomas- had in this respect; does the $\I{Alloc}$ rule in HeapLang
not introduce non-determinism in a similar way? Why is this not problematic?)
Lars' suggestion was to additionally parameterize the operational semantics by
an input-stream $\stdin$.
The updated \textsc{IOLoad} load rule might then look something like this:
\begin{mathpar}
  \inferrule[IOLoad]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO
  \\ n = \#\{i \mid t[i] = ( \X{IORead},\_\;,\_) \}  }
  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
    (r[\X{dst} := \stdin[n]], m, t \app (\X{IORead}, a, \stdin[n])) }
\end{mathpar}

When parameterizing our development with multiple values, we could either write
them out as different Coq \texttt{Parameters} where we need them, or store them
as fields in our \texttt{DriverG} typeclass (the latter is the approach taken
in e.g.\ the code-base of ~\cite{sammler-2020}).
\end{remark}

We now have two options when it comes to modeling the memory $m$:
\begin{enumerate}
\item Include the \MMIO addresses into the map. Notice that, for any address in \MMIO, the specific value held by the map $m$ for
that address is now irrelevant. Indeed, $m$ is never read nor modified for
addresses in \MMIO (possible exception: see next remark). One could choose to enforce that a particular dummy value is stored in $m$ for these addresses. Instead, we choose to leave them
unconstrained.
\item Have $m$ model the physical RAM addresses only. This model is closer to
  the conceptual model of state that we wish to implement, since it does not
  require us to talk about the values in the heap that \MMIO locations are
  associated with. In case it is not harder to implement (which I currently do not think
  it is), it is to be preferred.
\end{enumerate}
Both approaches require us to make sure that no points-to chunks $l \mapsto \_$
with $l \in \MMIO$ can be created. In the next section, we assume the second
approach (until someone proves me wrong about it working).

\begin{remark}
  Additionally, we have the option of disallowing an execution
  with a \X{PC} register pointing to a $\MMIO$ location in the semantics.
  First off, consider that this is an edge case; since we always have points-to
  chunks describing the layout of code in memory (and a points-to chunk for
  location $l$ doubles as a proof that $l \not\in \MMIO$), it is trivial to prove that
  this case would not occur in any of our current examples.
  If we make no changes to the current code base, this behavior would be
  allowed, and the following would happen in both cases described above:
  \begin{enumerate}
  \item An unconstrained value is read from the location. Any instruction could
    be executed.
  \item The predicate $\I{MemLocate}$ in $\I{lang.v}$ will make sure
  that the default value $0$ is read and decoded into an instruction when we try
  to execute an \MMIO location.
  This would not cause a conflict with the current WP rules,
  since a points-to chunk is required in the precondition, ensuring that this
  case does not occur.
  Another alternative is to read the encoding of a $\X{NOP}$ instruction, instead of
  just the value $0$.
\end{enumerate}
  If we were to simply disallow execution when the PC points into \MMIO, we would make the semantics of all instructions slightly more complicated, but in a
  uniform way.
  We could simply add $a \not\in \MMIO$ to the predicate $\I{isCorrectPC}$ (or
  add it as an additional predicate), which is used to define failing execution.
  It would again not be a drastic change, since a points-to for the
  address $a$ the PC points to is required by the WP rules regardless, from which we
  can prove that $a \not\in \MMIO$.
\end{remark}

\begin{remark} With this presentation, the specification of the
``machine'' is very much decoupled from the model of the devices it might be
communicating with. This is a good thing, I think.

Nevertheless, one could consider an alternative presentation where the model of
the devices is more tightly integrated with the semantics of the machine. For
instance, one could make the operational semantics parameterized with the
devices' model, where each device is modeled as having some internal state, the
ability to react on reads or writes, or to perform an internal step. Then, the
operational semantics would either step the usual way, or whenever a device
steps.

I believe this would be somewhat similar to the semantics of I/O system calls
through a foreign function interfaces as formalized in
CakeML~\cite{cakeml-vstte17io}, and also in Perennial~\cite{perennial-lang}.
%

\end{remark}

\begin{remark} Alix says that the proposed style of operational semantics
(using a trace) is close to the semantics of \texttt{volatile} as in
CompCert---which is also how memory mapped addresses seem to be exposed to C
compilers in practice. So this is probably a good sign. 
\end{remark}

@@

The simplest way to account for the trace is to directly expose it in a
monolithic fashion. In that case, the state interpretation additionally holds a
resource for the trace:

@@

\begin{remark}
For the case where the \MMIO locations are still part of the heap $m$, this
would look as follows:
\[
  \X{state\_interp} \; (r,m,t) :=
    \X{gen\_heap\_ctx}\; r \ast
    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
    {\color{BrickRed}\tracefull{t}}
\]

Sanity check: we never have allocation of new locations $l$ in our capability machine, but if we
did, that would work in both settings by requiring that $l \not\in \MMIO$ in the
operational semantics rule for \textsc{Alloc}.
\end{remark}

@@

The corresponding resource algebra is $\authm(\exm(\X{Trace}))$. A few relevant
rules are:

@@

\begin{remark} This is very coarse-grained: either one has the full
ownership for performing I/O and reasoning about it, or one cannot know anything
about it.

One application of this second extension could be the verification of an example
involving ``multiplexed'' I/O, where two separate parts of the code are granted
separate ownership over separate MMIO addresses. These two separate pieces of
code would be verified separately; then, in the end, one could prove that one
gets \emph{some} interleaving of all the events emitted by both components.
\end{remark}

@@

\begin{remark}
  As an alternative to having the set of MMIO addresses (\MMIO) as a global
  parameter, we might want to make it part of the state in the operational
  semantics. In that case, we would need another resource algebra to model the
  MMIO locations, i.e. add do the state interpretation we had before:
%
\[
  \X{state\_interp} \; (r,m,t,\MMIO) := \ldots \ast \MMIOag
\]

(where $\authm(\exm(\X{Trace}))$ is the resource algebra).
\end{remark}

\begin{remark}
In the future, we might be interested in the dynamic allocation of MMIO memory.
The question is what this would mean, though, since the set of available devices
allowing for MMIO access and the concrete buffers they provide will most likely
still be modeled as fixed at runtime.
Rather, this would allow us to take (un)mapped MMIO locations, connected to the different devices, and (un)map them anywhere in main memory.
This would require a different way of modeling MMIO, where we need both a notion
of the total pool of possible MMIO locations, and a description of the currently
mapped locations. We could model this using an authoritative RA.
\end{remark}

@@

The theorem then becomes:
%
\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \trInv \vdash \hoare{\TRUE}{e}{x.\; \operatorname{\Sf{lowval}} x} \\
    (C[e]);(\emptyset, \Sf{OK}); \emptyset \longrightarrow^* T';(h',g'); t
  }
  {g' = \Sf{OK} \wedge P(t)}
\end{mathpar}
, where we introduced the following notation:
%
\[
  \begin{array}{lcl}
    \trace(t) & \eqdef & \tracefrag{t}\\
    \trInv & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast \pure{\! P(t)}}
  \end{array}
\]

@@

Then, $\VR^{\notMMIO}$ is similar to the $\lowval$ predicate of OCPL.
Intuitively, a value in the relation does not directly point to memory-mapped
locations, nor can it gain access to memory-mapped locations indirectly,
similarly to how $\lowval$s cannot grant access to high locations
directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the registers
with values in $\VR^{\notMMIO}$, then the invariants in a private future world
of our starting world will be satisfied at the end of execution.

\begin{remark} As it is defined above, the ``generalized'' value relation
is in fact not very useful for instantiation of $\Psi$ other than $\notMMIO$.
Indeed, even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$
region, the value relation does not grant any corresponding resources. The fix
would be to use the generalized trace resources mentioned previously, and grant
ownership for the part of the trace corresponding to the addresses in
$[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of
addresses in \MMIO.
\end{remark}

@@

The code that is verified (similar to the expression $e$ in OCPL) is here given
as a set of \emph{entrypoints} (addresses) $A_{\X{entry}}$.
%
The capability $(\X{RX}, g, b, e, a_i)$ then points to the $i$th entry point of
the code we want to gradually verify.

@@

Notice the universal quantification over worlds in the precondition;
this needs to be there, since we do not know in what world our code will be
called.

\begin{remark}
  We could simplify the theorem statement by only considering the case of a
  single entrypoint. This is equivalent: a piece of code that wants to expose
  two distinct ``methods'' can implement a single toplevel entrypoint that
  immediately returns two pointers that can then be used to invoke the two
  methods.
  %
  This would make the toplevel theorem simpler, at the cost of somewhat more
  contrieved calling convention between trusted and untrusted code.
  Another alternative is to provide an opcode to specify the operation that
  needs to be performed.
  This is slightly less general, since it does not allow providing read access
  only to an adversary.
\end{remark}

\begin{remark}
The universal quantification on worlds was not explicitly present in the OCPL
formalization.
The reason for that is that (I think so at least, do not quote me
on this - Thomas) they
used the built-in Iris worlds within their definition of weakest precondition,
by leveraging the state interpretation, which is universally quantified over in
the definition of weakest precondition.
As we discussed before, if we do away
with local capabilities, it should be possible to fall back onto Iris' notion of
worlds, which would allow us to remove the universal quantification in the
above definition too because it would be implicitly present in the WP inside $\ER$.

It would be interesting to have a discussion about how to do away with the
worlds if we remove local capabilities, regardless of whether we will in the
end, since that might shed some light on why the worlds look the way they do
right now.
\end{remark}

\begin{remark}
  Note that we could have written
  $\ldots \vdash \forall W \ldotp \; \VR^{\notMMIO} (W) (\X{E}, \ldots)$
  in the above theorem, instead of using the expression relation. This is
  equivalent, and easily unfolds to the former statement through the definition
  of the value relation. I wrote $\ER$ to place the focus on the execution of
  the driver being safe. I consider this a matter of taste.
\end{remark}

@@

@@

Notice how the syntactic check for the registers is only performed on the set of
registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to
the driver address space (TODO: we should change this though and add a
requirement to $\initOKo$).
(TODO: introduce notation for trace-owndership and predicate P holding)

\begin{remark}
  If useful, the syntactic check can be relaxed to allow any capabilities that
  do not point into the MMIO region or the driver's memory. One would define the
  predicate $\notMapsToR{R}{w}$, that takes a word, memory region or register
  file and checks if it points into the forbidden region $R$ as follows:

Then, one would use $\notMapsToR{\MMIO \cup [b,e)}$ in place of $\nonCap{}$.
\end{remark}

@@

>>>>

\newcommand{\rk}[1]{\hyperref[{rk:#1}]{\textbf{[Rk~\ref*{rk:#1}]}}}

@@

Notice how in the second rule, we read an arbitrary integer $x$, and record it
in the trace~\rk{opsem:nondet}.
%
The \I{Store} rule would be similar.

\newcommand{\stdin}{\X{in}}

We now have two options when it comes to modeling the memory $m$:
\begin{enumerate}
  \item Include the \MMIO addresses into the map. Notice that, for any address
    in \MMIO, the specific value held by the map $m$ for that address is now
    irrelevant. Indeed, $m$ is never read nor modified for addresses in \MMIO
    (possible exception: see \rk{opsem:pc_in_mmio}). One could choose to enforce
    that a particular dummy value is stored in $m$ for these addresses or to
    leave them unconstrained.
%
  \item Have $m$ model the physical RAM addresses only. This model is closer to
    the conceptual model of state that we wish to implement, since it does not
    require us to talk about the values in the heap that \MMIO locations are
    associated with. It however probably requires threading through the
    operational semantics the invariant that $m$ is disjoint from \MMIO.
\end{enumerate}
%
Both approaches require us to make sure that no points-to chunks $l \mapsto \_$
with $l \in \MMIO$ can be created. In the next section, we assume the second
approach~\rk{opsem:pc_in_mmio}.

@@

The simplest way to account for the trace is to directly expose it in a
monolithic fashion. In that case, the state interpretation additionally holds a
resource for the trace~\rk{resources:mmio_in_m}:

@@

@@

The corresponding resource algebra is
$\authm(\exm(\X{Trace}))$~\rk{resources:fine_grained_trace}. A few relevant
rules are:

@@

To simplify notation, let us introduce the following definitions for user-level
ownership over the trace, and preservation of an invariant on the trace,
respectively:

\[
  \begin{array}{rcl}
    \trace(t) & \eqdef & \tracefrag{t} \\
    \trInv[P][\iota] & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast \pure{P(t)}}
  \end{array}
\]

\rk{resources:mmio_in_heap} \rk{resources:dynamic_mmio}

@@

@@

The theorem then becomes:
%
\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \trInv \vdash \hoare{\TRUE}{e}{x.\; \operatorname{\Sf{lowval}} x} \\
    (C[e]);(\emptyset, \Sf{OK}); \emptyset \longrightarrow^* T';(h',g'); t
  }
  {g' = \Sf{OK} \wedge P(t)}
\end{mathpar}

@@

Then, $\VR^{\notMMIO}$ is similar to the $\lowval$ predicate of OCPL.
Intuitively, a value in the relation does not directly point to memory-mapped
locations, nor can it gain access to memory-mapped locations indirectly,
similarly to how $\lowval$s cannot grant access to high locations
directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the registers
with values in $\VR^{\notMMIO}$, then the invariants in a private future world
of our starting world will be satisfied at the end of execution.
%
\rk{toplevel:v_psi}

@@

The code that is verified (similar to the expression $e$ in OCPL) is here given
as a set of \emph{entrypoints} (addresses) $A_{\X{entry}}$.
%
The capability $(\X{RX}, g, b, e, a_i)$ then points to the $i$th entry point of
the code we want to gradually verify.
%
\rk{toplevel:single_entrypoint}

@@

Notice the universal quantification over worlds in the precondition;
this needs to be there, since we do not know in what world our code will be
called.
%
\rk{toplevel:ocpl_explicit_worlds}

@@

\rk{toplevel:less_restricted_syntactic_check_adverary}

@@

Notice how the syntactic check for the registers is only performed on the set of
registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to
the driver address space (TODO: we should change this though and add a
requirement to $\initOKo$).

@@

\section{Remarks}

\subsection{Operational Semantics}

\subsubsection{Non-determinism}
\label{rk:opsem:nondet}

With this presentation, the specification of the ``machine'' is very much
decoupled from the model of the devices it might be communicating with. This is
a good thing, I think.

One possible downside is that this makes the operational semantics
non-deterministic, which might or might not be an issue in the future.

There are several ways in which the operational semantics could be made ``more
deterministic''.

One could additionally parameterize the operational semantics by an input-stream
$\stdin$. The updated \textsc{IOLoad} load rule might then look something like
this:
\begin{mathpar}
  \inferrule[IOLoad]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO
  \\ n = \#\{i \mid t[i] = ( \X{IORead},\_\;,\_) \}  }
  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
    (r[\X{dst} := \stdin[n]], m, t \app (\X{IORead}, a, \stdin[n])) }
\end{mathpar}

When parameterizing our development with multiple values, we could either write
them out as different Coq \texttt{Parameters} where we need them, or store them
as fields in a \texttt{DriverG} typeclass (the latter is the approach taken in
e.g.\ the code-base of~\cite{sammler-2020}).

One could also take as a parameter a predicate that restrict the set of traces
that are allowed (that is what~\cite{sammler-2020} do).

We could also express more precisely the devices' model, and equip each device
with some internal state, the ability to react on reads or writes, or to perform
an internal step. Then, the operational semantics would either step the usual
way, or whenever a device steps.

I believe this would be somewhat similar to the semantics of I/O system calls
through a foreign function interfaces as formalized in
CakeML~\cite{cakeml-vstte17io}, and also in Perennial~\cite{perennial-lang}.

\subsubsection{PC in MMIO}
\label{rk:opsem:pc_in_mmio}

Additionally, we have the option of disallowing an execution with a \X{PC}
register pointing to a $\MMIO$ location in the semantics.
%
First off, consider that this is an edge case; since we always have points-to
chunks describing the layout of code in memory (and a points-to chunk for
location $l$ doubles as a proof that $l \not\in \MMIO$), it is trivial to prove
that this case would not occur in any of our current examples.
%
If we make no changes to the current code base, this behavior would be allowed,
and the following would happen in both cases described above:
%
\begin{enumerate}
  \item An unconstrained value is read from the location. Any instruction could
    be executed.
%
  \item The predicate $\I{MemLocate}$ in $\I{lang.v}$ will make sure that the
    default value (e.g. $0$) is read and decoded into an instruction when we try
    to execute an \MMIO location. This would not cause a conflict with the
    current WP rules, since a points-to chunk is required in the precondition,
    ensuring that this case does not occur.
    %
    Another alternative is to read the encoding of a $\X{NOP}$ instruction,
    instead of just the value $0$.
\end{enumerate}
%
If we were to simply disallow execution when the PC points into \MMIO, we would
make the semantics of all instructions slightly more complicated, but in a
uniform way.
%
We could simply add $a \not\in \MMIO$ to the predicate $\I{isCorrectPC}$ (or add
it as an additional predicate), which is used to define failing execution. It
would again not be a drastic change, since a points-to for the address $a$ the
PC points to is required by the WP rules regardless, from which we can prove
that $a \not\in \MMIO$.

\subsection{Resources}

\subsubsection{MMIO overlapping with $m$}
\label{rk:resources:mmio_in_m}

For the case where the \MMIO locations are still part of the heap $m$, this
would look as follows:
\[
  \X{state\_interp} \; (r,m,t) :=
    \X{gen\_heap\_ctx}\; r \ast
    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
    {\color{BrickRed}\tracefull{t}}
\]

Sanity check: we never have allocation of new locations $l$ in our capability
machine, but if we did, that would work in both settings by requiring that
$l \not\in \MMIO$ in the operational semantics rule for \textsc{Alloc}.

\subsubsection{Fine-grained trace resources}
\label{rk:resources:fine_grained_trace}

This is very coarse-grained: either one has the full
ownership for performing I/O and reasoning about it, or one cannot know anything
about it.

One application of this second extension could be the verification of an example
involving ``multiplexed'' I/O, where two separate parts of the code are granted
separate ownership over separate MMIO addresses. These two separate pieces of
code would be verified separately; then, in the end, one could prove that one
gets \emph{some} interleaving of all the events emitted by both components.

\subsubsection{MMIO in the heap}
\label{rk:resources:mmio_in_heap}

As an alternative to having the set of MMIO addresses (\MMIO) as a global
parameter, we could make it part of the state in the operational semantics. In
that case, we would need another resource algebra to model the MMIO locations,
i.e. add do the state interpretation we had before:
%
\[
  \X{state\_interp} \; (r,m,t,\MMIO) := \ldots \ast \MMIOag
\]

(where $\authm(\exm(\X{Trace}))$ is the resource algebra).

\subsubsection{Dynamic MMIO}
\label{rk:resources:dynamic_mmio}

In the future, we might be interested in the dynamic allocation of MMIO memory.
The question is what this would mean, though, since the set of available devices
allowing for MMIO access and the concrete buffers they provide will most likely
still be modeled as fixed at runtime.
Rather, this would allow us to take (un)mapped MMIO locations, connected to the different devices, and (un)map them anywhere in main memory.
This would require a different way of modeling MMIO, where we need both a notion
of the total pool of possible MMIO locations, and a description of the currently
mapped locations. We could model this using an authoritative RA.

\subsection{Toplevel theorem}

\subsubsection{$\VR^\Psi$ with arbitrary $\Psi$}
\label{rk:toplevel:v_psi}

As it is defined above, the ``generalized'' value relation
is in fact not very useful for instantiation of $\Psi$ other than $\notMMIO$.
Indeed, even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$
region, the value relation does not grant any corresponding resources. The fix
would be to use the generalized trace resources mentioned previously, and grant
ownership for the part of the trace corresponding to the addresses in
$[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of
addresses in \MMIO.

\subsubsection{Driver with single entrypoint}
\label{rk:toplevel:single_entrypoint}

We could simplify the theorem statement by only considering the case of a
single entrypoint. This is equivalent: a piece of code that wants to expose
two distinct ``methods'' can implement a single toplevel entrypoint that
immediately returns two pointers that can then be used to invoke the two
methods.
%
This would make the toplevel theorem simpler, at the cost of somewhat more
contrieved calling convention between trusted and untrusted code.
Another alternative is to provide an opcode to specify the operation that
needs to be performed.
This is slightly less general, since it does not allow providing read access
only to an adversary.

\subsubsection{OCPL and explicit worlds}
\label{rk:toplevel:ocpl_explicit_worlds}

The universal quantification on worlds was not explicitly present in the OCPL
formalization.
The reason for that is that (I think so at least, do not quote me
on this - Thomas) they
used the built-in Iris worlds within their definition of weakest precondition,
by leveraging the state interpretation, which is universally quantified over in
the definition of weakest precondition.
As we discussed before, if we do away
with local capabilities, it should be possible to fall back onto Iris' notion of
worlds, which would allow us to remove the universal quantification in the
above definition too because it would be implicitly present in the WP inside $\ER$.

It would be interesting to have a discussion about how to do away with the
worlds if we remove local capabilities, regardless of whether we will in the
end, since that might shed some light on why the worlds look the way they do
right now.

\subsubsection{Less restricted but still syntactic check for the adversary memory}
\label{rk:toplevel:less_restricted_syntactic_check_adverary}

If useful, the syntactic check can be relaxed to allow any capabilities that
do not point into the MMIO region or the driver's memory. One would define the
predicate $\notMapsToR{R}{w}$, that takes a word, memory region or register
file and checks if it points into the forbidden region $R$ as follows:

Then, one would use $\notMapsToR{\MMIO \cup [b,e)}$ in place of $\nonCap{}$.

%% Biblio
